Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-22 main.tex
fonet->MLL

\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\rsubstof}[1]{\rho_{#1}}
\newcommand{\dsubstof}[1]{\delta_{#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}
\newcommand{\labelof}[1]{\ell(#1)}
\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

@@

More formally, let $\Gamma$ be a sequent with $\graphof\Gamma=\gC$, to which we
exhaustively apply the following
subformula rewriting steps, to obtain a sequent $\frameof\Gamma$:
\begin{enumerate}
\item {\bf Encode dependencies as fresh links.} For each dependency
  $\pair{\single x,\single y}$ in $\gC$, with corresponding
  subformulas $\exists x A$ and $\forall y B$ in $\Gamma$, we pick a
  fresh (nullary) predicate symbol $q$, and then replace $\exists
  x. A$ by $q \cand \exists x. A$, and replace $\forall y. B$ by
  $\dual q \cor \forall y. B$.
\item {\bf Erase quantifiers.} After step 1, remove all the
  quantifiers, i.e., replace $\exists x.A$ by $A$ and replace $\forall
  y.B$ by $B$ everywhere.
\item {\bf Simplify atoms.} After step 2, replace every predicate
  $p(t_1 \cdots t_n)$ (resp. $\dual p(t_1 \cdots t_n)$) with a nullary
  predicate symbol $p$ (resp.~$\dual p$)
\end{enumerate}
The $\linkingof{\frameof\Gamma}$ consists of the pairs induced by
$\linkingof\gC$ and the new pairs $\set{q,\dual q}$ introduced in
step~1 above.  Now we define the \bfit{frame} of $\gC$, denoted
$\frameof\gC$, as
$\tuple{\graphof{\frameof\Gamma},\linkingof{\frameof\Gamma}}$.

@@

Note that the linking $\lambda_m$ is a valid $\MLLm$ proof net.

We define the \bfit{frame} of a linked fograph to be the linked cograph defined as follows:

>>>>

\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\rsubstof}[1]{\rho_{#1}}
\newcommand{\dsubstof}[1]{\delta_{#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\linking}{\sim}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}
\newcommand{\labelof}[1]{\ell(#1)}
\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

@@

More formally, let $\Gamma$ be a sequent with $\graphof\Gamma=\gC$, to which we
exhaustively apply the following
subformula rewriting steps, to obtain a sequent $\frameof\Gamma$:
\begin{enumerate}
\item {\bf Encode dependencies as fresh links.} For each dependency
  $\pair{\single x,\single y}$ in $\gC$, with corresponding
  subformulas $\exists x A$ and $\forall y B$ in $\Gamma$, we pick a
  fresh (nullary) predicate symbol $q$, and then replace $\exists
  x. A$ by $q \cand \exists x. A$, and replace $\forall y. B$ by
  $\dual q \cor \forall y. B$.
\item {\bf Erase quantifiers.} After step 1, remove all the
  quantifiers, i.e., replace $\exists x.A$ by $A$ and replace $\forall
  y.B$ by $B$ everywhere.
\item {\bf Simplify atoms.} After step 2, replace every predicate
  $p(t_1 \cdots t_n)$ (resp. $\dual p(t_1 \cdots t_n)$) with a nullary
  predicate symbol $p$ (resp.~$\dual p$)
\end{enumerate}
The $\linkingof{\frameof\Gamma}$ consists of the pairs induced by
$\linkingof\gC$ and the new pairs $\set{q,\dual q}$ introduced in
step~1 above. We call $\frameof\Gamma$ the \bfit{frame} of $\Gamma$ and we define the \bfit{frame} of $\gC$, denoted
$\frameof\gC$, as
$\tuple{\graphof{\frameof\Gamma},\linkingof{\frameof\Gamma}}$, and we immediately have the following:

@@

Let $\Gamma$ be a propositional sequent and $\linkingof\Gamma$ be a
linking for $\graphof\Gamma$. A conjunction formula $A\vlan B$ is
\bfit{splitting} or a \bfit{splitting tensor} if
$\Gamma=\Gamma',A\vlan B,\Gamma''$ and
$\mathord{\linkingof\Gamma}=\mathord{\linking_1}\cup\linking_2$, such
that $\linking_1$ is a linking for $\graphof{\Gamma',A}$ and
$\linking_2$ is a linking for $\graphof{B,\Gamma''}$, i.e., removing
the $\vlan$ from $A\vlan B$ splits the linked fograph $\tuple{
  \graphof\Gamma,\linkingof\Gamma}$ into two fographs.
We say that $\tuple{
  \graphof\Gamma,\linkingof\Gamma}$ is \bfit{mixed} iff 
$\Gamma=\Gamma',\Gamma''$ and
$\mathord{\linkingof\Gamma}=\mathord{\linking_1}\cup\linking_2$, such
that $\linking_1$ is a linking for $\graphof{\Gamma'}$ and
$\linking_2$ is a linking for $\graphof{\Gamma''}$.
Finally, $\tuple{
  \graphof\Gamma,\linkingof\Gamma}$ is \bfit{splittable} if it is mixed or has a splitting tensor.

The purpose of introducing the frame is the following theorem.

\begin{thm}
  \label{thm:splitting}
  Let $\Gamma$ be a propositional sequent containing only atoms
  and $\vlan$-formulas, and $\linkingof\Gamma$ be a
  linking for $\graphof\Gamma$. If $\tuple{
    \graphof\Gamma,\linkingof\Gamma}$ does not induce a bimatching then it is splittable.
\end{thm}

This is the well-know splitting-tensor-theorem
\cite{girard:87,danos:regnier:89}, adapted for the presence of
$\mix$. In the setting of linked cographs, it has first been proved by
Retor\'e~\cite{retore:03,retore:??}\lutz{that ``handsome'' report he mentioned in
  the talk}. We use it now for our sequentialization:

\begin{thm}
  \label{thm:fonet->MLL}
  Let $\tuple{\gC,\linkingof\gC}$ be a fonet, and let $\Gamma$ be a sequent with $\graphof\Gamma=\gC$. Then there is an
  $\FOMLL$-proof $\Pi$ of $\Gamma$, such that
  $\fographof\Pi=\tuple{\gC,\linkingof\gC}$.
\end{thm}

\begin{proof}
  Let $\dsubstof\gC$ be the dualizer of $\tuple{\gC,\linkingof\gC}$.  We
  proceed by induction on the size of $\Gamma$ (i.e., the number of
  symbols in it, without counting the commas). If $\Gamma$ contains a
  formula with $\vlor$-root, or a formula $\forall x.A$, we can
  immediately apply the $\vlor$-rule or the $\forall$-rule of $\FOMLL$
  and proceed by induction hypothesis. If $\Gamma$ contains a formula
  $\exists x.A$ such that the correponding binder $\single x$ in $\gC$
  has no dependency, then we can apply the $\exists$-rule, chosing the
  term $t$ as determined by $\dsubstof\gC$, and proceed by induction
  hypothesis.  Hence, we can now assume that $\Gamma$ contains only
  atoms, $\vlan$-formulas, or formulas of shape $\exists x.A$, where
  the vertex $\single x$ has dependencies. Then the frame
  $\tuple{\graphof{\frameof\Gamma},\linkingof{\frameof\Gamma}}$ does
  not induce a bimatching and contains only atoms and
  $\vlan$-formulas, and is therefore splittable. If it is mixed, then
  we can apply the $\mix$-rule to $\Gamma$ and apply the induction
  hypothesis to the two components. If it is not mixed then there must
  be a splitting tensor. If the splitting $\vlan$ is already in
  $\Gamma$, then we can apply the $\vlan$-rule and proceed by
  induction hypothesis on the two branches. However, if
  $\frameof\Gamma$ is not mixed and all splitting tensors are
  $\vlan$-formulas introduced in step~1 of the frame construction,
  then we get a contradiction as in that case there must be a $\vlor$- or $\forall$-formula in $\Gamma$.
  \lutz{can anyone give a good argument here?}
\end{proof}
